Describing demographic models is hard

### 2020-05-20 paper.tex
sharpen focus of Martin section and move material to conclusions

@@

\section{A missing parameter}

In another publication using this model~\citep{martin2017human},
a separate error was introduced: the model itself was defined as suggested
in the documentation (using updated parameters
from~\citet{gravel2011demographic}) and inspected
using the \msprime\ debugging tools.
After these initial checks were made, however, the simulation
was performed without passing the parameter of demographic events,
so that the three populations remained separated with low levels of migration and
never merged as expected (Fig~\ref{fig:prs}A),
leading to a vast overestimate of the divergence across human populations.
Whereas the correct model predicts a mean $F_{ST}$ of
$0.05 - 0.10$ across the three population, the simulated model generated $F_{ST}$
ranging between $0.3 - 0.6$, depending on the populations considered.
Overall diversity was also strongly affected: expected heterozygosity was more than
doubled in African populations but just $30-40\%$ the expected
levels in Eurasian populations when compared to expectations from the correct model.

This simulation was performed to assess the
transferability of polygenic risk scores across human populations, in particular to
explore how human demographic history and population structure affects our ability
to predict genetic risk in populations of different ancestry than the originally studied
population.
The excess divergence exaggerated the role that demography plays
in limiting the transferability of genetic risk scores across populations.
The resulting publication has been influential in the discussion of
health inequalities and genomics, with over 350 citations since 2017.
We note that while difficulties in transferability remain in the corrected model
(Fig.~\ref{fig:prs}), risk prediction in each population is significantly improved
when using the correct demographic model
(compare to Fig.~5 in~\citet{martin2017human}).

These errors suggests three lessons.
First, the design of user interface and API for scientific software matters,
and many bugs can be prevented by using more intuitive interfaces.
Whereas the original \msprime\ required the user to pass
three separate parameters to specify a demographic model, \msprime~ver.~1.0
will introduce a Demography class, which wraps these three parameters.
Thus, instead of writing
\begin{lstlisting}[frame=single]
dbg = msprime.DemographyDebugger(
  population_configurations=population_configurations,
  migration_matrix=migration_matrix,
  demographic_events=demographic_events)
dbg.print_history()
ts = msprime.simulate(
  population_configurations=population_configurations,
  migration_matrix=migration_matrix,
  demographic_events=demographic_events)
\end{lstlisting}
with the error-inducing need to re-enter parameter information,
we would now write
\begin{lstlisting}[frame=single]
demography = msprime.Demography(
  population_configurations=population_configurations,
  migration_matrix=migration_matrix,
  demographic_events=demographic_events)
dbg = demography.debug()
dbg.print_history()
ts = msprime.simulate(demography=demography)
\end{lstlisting}

Second, testing of the final simulated data is important.
This can be challenging, because the amount of simulated data can be large.
However, recent progress in computing summary statistics from tree
sequence data can make this easier~\citep{ralph2020efficiently}.
Even if the entire simulation data cannot be easily evaluated, subsets of the data
can be examined to identify large errors.

@@

Finally, from a genetics perspective, the corrected simulations indicate that
the accuracy of genetic risk scores is still substantially reduced in understudied
populations (Fig.~\ref{fig:prs}E-G), supporting the main conclusion of~\citet{martin2017human}.
However, the reduction is much less pronounced than reported.
In particular, we do not observe large differences in mean risk prediction
across populations (Fig.~\ref{fig:prs}C,D), as presented in~\citet{martin2017human}.
At least for the neutral polygenic architectures considered here, genetic drift alone
does not appear to induce large biases in mean predicted risk across ethnicities.

The implementation of complex demographic models is error prone, and such errors
can have a large impact on downstream analysis and interpretation.
Recognition of these difficulties drove efforts to peer-review demographic model
implementation in \stdpopsim~\citep{adrion2019community}, with quality-controlled
models and simulation resources for a growing number of commonly studied species.

We therefore recommend the following steps to ensure more robust simulations.
First, we recommend that demographic models used in publications
are verified by a second pair of eyes in code review or an independent implementation.
If the implementation is complex enough that code review is impractical, the risk
that bugs are introduced is extremely high and additional caution and statistical
verification is warranted.
Second, we recommend basic statistical verification of all simulated data.
If simulations are so large that statistical validation is burdensome,
subsets of the simulated data can be analyzed~\citep{ralph2020efficiently}.
Such tests would have caught the error in~\citet{martin2017human}.

>>>>

\aprcomment{title suggestions:\\
}

@@

\section{A missing parameter (or) }

In another publication using this model~\citep{martin2017human},
a separate error was introduced: the model itself was defined as suggested
in the documentation (using updated parameters
from~\citet{gravel2011demographic}) and inspected
using the \msprime\ debugging tools.
After these initial checks were made, however, the simulation
was performed without passing the parameter of demographic events,
so that the three populations remained separated with low levels of migration and
never merged as expected (Fig~\ref{fig:prs}A),
leading to a vast overestimate of the divergence across human populations.
While the correct model predicts a mean $F_{ST}$ of
$0.05 - 0.10$ across the three population, the simulated model generated $F_{ST}$
ranging between $0.3 - 0.6$, depending on the populations considered.
Overall diversity was also strongly affected: expected heterozygosity was more than
doubled in African populations but just $30-40\%$ the expected
levels in Eurasian populations when compared to expectations from the correct model.

This simulation was performed to assess the
transferability of polygenic risk scores across human populations, in particular to
explore how human demographic history and population structure affect our ability
to predict genetic risk in populations of different ancestry than the originally studied
population.
The resulting publication has been influential in the discussion of
health inequalities and genomics, with over 350 citations since 2017.
But while the first implementation error had a small effect on resulting diversity statistics,
this error more strongly impacts simulated diversity and interpretations based on that data.
The excess divergence under the incorrect model here exaggerated the role that demography
plays in limiting the transferability of genetic risk scores across populations.

Difficulties in transferability remain in the corrected model (Fig.~\ref{fig:prs}), although
risk prediction in each population is significantly improved when using the correct
demographic model (compare to Fig.~5 in~\citet{martin2017human}).
The corrected simulations indicate that the accuracy of genetic risk scores is still
substantially reduced in understudied populations (Fig.~\ref{fig:prs}E-G),
supporting the main conclusion of~\citet{martin2017human}.
However, the reduction is much less pronounced than reported.
In particular, we do not observe large differences in mean risk prediction
across populations (Fig.~\ref{fig:prs}C,D), as presented in~\citet{martin2017human}.
At least for the neutral polygenic architectures considered here, genetic drift alone
does not appear to induce large directional biases in mean predicted risk across ethnicities.

The implementation of complex demographic models is error prone, and such errors
can have a large impact on downstream analysis and interpretation. 
The discovery and correction of the simulation errors discussed here underscore
how API design choices and 
and we recommend the following steps to ensure more robust simulations.

First, the design of user interface and API for scientific software matters,
and many bugs can be prevented by using more intuitive interfaces.
Whereas the original \msprime\ required the user to pass
three separate parameters to specify a demographic model, \msprime~ver.~1.0
will introduce a Demography class, which wraps these three parameters.
Thus, instead of writing
\begin{lstlisting}[frame=single]
dbg = msprime.DemographyDebugger(
  population_configurations=population_configurations,
  migration_matrix=migration_matrix,
  demographic_events=demographic_events)
dbg.print_history()
ts = msprime.simulate(
  population_configurations=population_configurations,
  migration_matrix=migration_matrix,
  demographic_events=demographic_events)
\end{lstlisting}
with the error-inducing need to re-enter parameter information,
we would now write
\begin{lstlisting}[frame=single]
demography = msprime.Demography(
  population_configurations=population_configurations,
  migration_matrix=migration_matrix,
  demographic_events=demographic_events)
dbg = demography.debug()
dbg.print_history()
ts = msprime.simulate(demography=demography)
\end{lstlisting}

Recognition of these difficulties drove efforts to peer-review demographic model
implementation in \stdpopsim~\citep{adrion2019community}, with quality-controlled
models and simulation resources for a growing number of commonly studied species.

We therefore recommend the following steps to ensure more robust simulations.
First, we recommend that demographic models used in publications
are verified by a second pair of eyes in code review or an independent implementation.
If the implementation is complex enough that code review is impractical, the risk
that bugs are introduced is extremely high and additional caution and statistical
verification is warranted.

Second, we recommend basic statistical verification of all simulated data.
If simulations are so large that statistical validation is burdensome,
subsets of the simulated data can be analyzed~\citep{ralph2020efficiently}.
Such tests would have caught the error in~\citet{martin2017human}.

----

These errors suggests three lessons.

Second, testing of the final simulated data is important. \aprcomment{verification of simulations, wrapped together}
This can be challenging, because the amount of simulated data can be large.
However, recent progress in computing summary statistics from tree
sequence data can make this easier~\citep{ralph2020efficiently}.
Even if the entire simulation data cannot be easily evaluated, subsets of the data
can be examined to identify large errors.

@@
